BayesPy: Variational Bayesian Inference in Python

### 2014-08-14 bayespy.tex
Minor fine tuning

    \plate {} {(yn)} {$n=1,\ldots,10$};
  \end{tikzpicture}
  \caption{Graphical model of the example problem.}
  \label{fig:graphical_model}
\end{figure}
In order to demonstrate BayesPy, this section solves an extremely simple problem
but which includes the main steps of using BayesPy.  The task is to estimate the
unknown mean and precision parameters of a Gaussian distribution given ten
observations.  Thus, the likelihood function is Gaussian:
\begin{align*}
  p(y_1,\ldots,y_{10}|\mu,\tau) &= \prod^{10}_{n=1}\mathcal{N}(y_n | \mu, \tau),
\end{align*}
where the Gaussian distribution is parameterized by precision instead of
variance.  The unknown mean $\mu$ and precision $\tau$ are given broad Gaussian
and gamma priors, respectively:
\begin{align*}
  p(\mu) &= \mathcal{N}(\mu|0, 10^{-6}) & p(\tau) &= \mathcal{G}(\tau|10^{-6},
  10^{-6})
\end{align*}
Figure~\ref{fig:graphical_model} shows the graphical model of this simple model.
The model is constructed in BayesPy by creating nodes:
\begin{lstlisting}
from bayespy.nodes import GaussianARD, Gamma
mu = GaussianARD(0, 1e-6)
tau = Gamma(1e-6, 1e-6)
\end{lstlisting}
These two nodes represent the two unknown variables that we are interested in.
They are the parameters of the Gaussian distribution of the observed variable:
\begin{lstlisting}
y = GaussianARD(mu, tau, plates=(10,))
\end{lstlisting}
The number of observations is given as plates which define the number of
repetitions.  Note that the syntax of the model construction is similar to the
mathematical formulation which defines the conditional distributions for each
variable.  Thus, it is easy to understand the model from the Python code.  After
the model has been constructed, some nodes are marked as observed and given
data:
\begin{lstlisting}
y.observe([4.5, 3.9, 6.3, 5.6, 4.9, 2.8, 7.4, 6.1, 4.8, 2.1])
\end{lstlisting}
Next, we want to find the posterior approximation for our latent variables.  We
create the VB inference engine:
\begin{lstlisting}
from bayespy.inference import VB
Q = VB(y, mu, tau)
\end{lstlisting}
The inference engine takes all the stochastic nodes of the model as input.  The
message passing algorithm can be run by calling \texttt{update} method:
\begin{lstlisting}
Q.update(repeat=100)
\end{lstlisting}
The algorithm is run for 100 iterations or until it converges.  In this case,
the algorithm converges in four iterations.  The posterior approximation can be
examined, for instance, by plotting the probability density functions of the
latent variables:
\begin{lstlisting}
import bayespy.plot as bpplt
import numpy as np
bpplt.pyplot.figure()
bpplt.pdf(Q[mu], np.linspace(0, 10, num=300))
bpplt.pyplot.figure()
bpplt.pdf(Q[tau], np.linspace(0, 2, num=300))
bpplt.pyplot.show()
\end{lstlisting}
Note that the standard \texttt{pyplot} module of Matplotlib is available at the
\texttt{plot} module of BayesPy.  Figure~\ref{fig:posterior} shows the resulting
approximate posterior distributions.
\begin{figure}
  \begin{center}
    \begin{subfigure}[b]{0.4\textwidth}
      \includegraphics[width=\textwidth]{fig_mu}
      \caption{$q(\mu)$}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
      \includegraphics[width=\textwidth]{fig_tau}
      \caption{$q(\tau)$}
    \end{subfigure}
  \end{center}
  \caption{The posterior approximations of $\mu$ and $\tau$.}
  \label{fig:posterior}
\end{figure}

>>>>

    \plate {} {(yn)} {$n=1,\ldots,10$};
  \end{tikzpicture}
  \caption{Graphical model of the example problem.}
  \label{fig:graphical_model}
\end{figure}
In order to demonstrate BayesPy, this section solves an extremely simple problem
but which includes the main steps of using BayesPy.  The task is to estimate the
unknown mean and precision parameters of a Gaussian distribution given ten
observations.  Thus, the likelihood function is Gaussian:
\begin{align*}
  p(y_1,\ldots,y_{10}|\mu,\tau) &= \prod^{10}_{n=1}\mathcal{N}(y_n | \mu, \tau),
\end{align*}
where the Gaussian distribution is parameterized by precision instead of
variance.  The unknown mean $\mu$ and precision $\tau$ are given broad Gaussian
and gamma priors, respectively:
\begin{align*}
  p(\mu) &= \mathcal{N}(\mu|0, 10^{-6}) & p(\tau) &= \mathcal{G}(\tau|10^{-6},
  10^{-6})
\end{align*}
Figure~\ref{fig:graphical_model} shows the graphical model of this simple model.
The model is constructed in BayesPy by creating nodes:
\begin{lstlisting}
from bayespy.nodes import GaussianARD, Gamma
mu = GaussianARD(0, 1e-6)
tau = Gamma(1e-6, 1e-6)
\end{lstlisting}
These two nodes represent the two unknown variables that we are interested in.
They are the parameters of the Gaussian distribution of the observed variable:
\begin{lstlisting}
y = GaussianARD(mu, tau, plates=(10,))
\end{lstlisting}
The number of observations is given as plates which define the number of
repetitions.  Note that the syntax of the model construction is similar to the
mathematical formulation which defines the conditional distributions for each
variable.  Thus, it is easy to understand the model from the Python code.  After
the model has been constructed, some nodes are marked as observed and given
data:
\begin{lstlisting}
y.observe([4.5, 3.9, 6.3, 5.6, 4.9, 2.8, 7.4, 6.1, 4.8, 2.1])
\end{lstlisting}
Next, we want to find the posterior approximation for our latent variables.  We
create the VB inference engine:
\begin{lstlisting}
from bayespy.inference import VB
Q = VB(y, mu, tau)
\end{lstlisting}
The inference engine takes all the stochastic nodes of the model as input.  The
message passing algorithm can be run by calling \texttt{update} method:
\begin{lstlisting}
Q.update(repeat=100)
\end{lstlisting}
The algorithm is run for 100 iterations or until it converges.  In this case,
the algorithm converges in four iterations.  The posterior approximation can be
examined, for instance, by plotting the probability density functions of the
latent variables:
\begin{lstlisting}
import bayespy.plot as bpplt
import numpy as np
bpplt.pyplot.figure()
bpplt.pdf(Q[mu], np.linspace(0, 10, num=300))
bpplt.pyplot.figure()
bpplt.pdf(Q[tau], np.linspace(0, 2, num=300))
bpplt.pyplot.show()
\end{lstlisting}
Figure~\ref{fig:posterior} shows the resulting approximate posterior
distributions.  The module \texttt{bayespy.plot} contains functions for simple
plotting tasks, such as plotting the probability density function or the Hinton
diagram of a variable.  Note that the standard \texttt{pyplot} module of
Matplotlib is available at the \texttt{plot} module of BayesPy.
\begin{figure}
  \begin{center}
    \begin{subfigure}[b]{0.4\textwidth}
      \includegraphics[width=\textwidth]{fig_mu}
      \caption{$q(\mu)$}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
      \includegraphics[width=\textwidth]{fig_tau}
      \caption{$q(\tau)$}
    \end{subfigure}
  \end{center}
  \caption{The posterior approximations of $\mu$ and $\tau$.}
  \label{fig:posterior}
\end{figure}
